Kindly Bent to Free Us

### 2019-07-10 kindly/lang.tex
kinds

\lang is a minimal ML-like language with
let-polymorphism, abstract types, and pairs. Its type system manages
linearity and borrows using kinds and (kind-) constrained types.
For ease of presentation, we keep the language as simple as possible.
\begin{itemize}
\item We showcase pattern matching on pairs rather than algebraic
  datatypes.
\item We have separate operators for borrowing and reborrowing (taking
  the borrow of a borrow).
\item We assume that regions are explicit in the code. (But we present
  an algorithm to infer them in \cref{regionannot}.)
\item We use de Bruijn-indexing to identify regions instead of region
  variables. 
\end{itemize}

The rest of this section formalizes  \lang: the syntax (\cref{syntax}),
the statics in terms of syntax directed typing (\cref{sdtyping}),
and a dynamics that is linearity-aware (\cref{sem}).

@@

We now present the essential parts of our \lang's type system. A complete
description of the system is available in \cref{appendix:sdtyping}.
In this section, we focus on chosen parts of our system
through a set of selected rules.
%
We introduce the following typing, kinding and constraint judgements:
\begin{description}
\item[$\inferS{C}{\E}{e}{\tau}$ --]
  Expression $e$ has type $\tau$ in environment $\E$ under constraints $C$.
\item[$\inferSK{C}{\E}{\tau}{k}$ --]
  Type $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
\item[$\entail{D}{C}$ --] Constraint $D$ implies constraint $C$.
\end{description}

@@

One novel feature or \affe is the use of kinds and qualified types to encode
linear and affine types. Before explaining the details of our type system, we
give a quick presentation of the kinds and constraints.
A kind $k$ can be either a kind variable, $\kvar$, or a constant.
Each constant is composed of a ``quality'' and a ``level''.
The quality describes the use pattern of the type and can be
either unrestricted $\kun$, affine $\kaff$ or linear $\klin$. The level
describes the nested regions in which it can be used. Levels are represented
by a natural number of $\infty$.
$0$ represents the normal scope outside of any region and is often omitted
(for instance $\kaff_0$ is simply noted $\kaff$).
$\infty$ is the empty region that is infinitely nested and is mostly used
as an upper bound.
For instance, the constraint $\Cleq{\kvar}{\kun_\infty}$ indicates that
$\kvar$ must be unrestricted, but can be local to a region.
%
Just as presented before, constraints present in schemes and type signatures
are simply a list of atomic constraints, which are composed
of kind inequalities.
%
Our inequality constraints are based on an inequality relation on
kind constants noted $\lk_\Lat$. This relation forms a lattice
and  is summarized in
\cref{sdtyp:lattice}. We have the basic inequalities on qualities:
$\kun_n \lk_\Lat \kaff_n \lk_\Lat \klin_n$ for arbitrary regions $n$.
Additionally, the inequality follows the region nesting, for
instance $\kaff_n \lk_\Lat \kaff_{n+1}$.

\begin{figure*}[!h]
  \input{infer/lattice}
  \caption{Lattice inequalities -- $k \lk_\Lat k'$}
  \label{sdtyp:lattice}
\end{figure*}

>>>>

\lang is a minimal ML-like language with
let-polymorphism, abstract types, and pairs. Its type system manages
linearity and borrows using kinds and (kind-) constrained types.
For ease of presentation, we keep the language as simple as possible.
\begin{itemize}
\item Pattern matching is demonstrated on pairs rather than algebraic
  datatypes.
\item There are separate operators for borrowing and reborrowing (taking
  the borrow of a borrow).
\item Regions are assumed to be explicit in the code. (But we present
  an algorithm to infer them in \cref{regionannot}.)
\item Regions are identified using de Bruijn-indexing instead of region
  variables. 
\end{itemize}

The rest of this section formalizes  \lang: the syntax (\cref{syntax}),
the statics in terms of syntax directed typing (\cref{sdtyping}),
and a dynamics that is linearity- and resource-aware (\cref{sem}).

@@

In this section, we focus on the essential and novel parts of the type system. A complete
description is available in \cref{appendix:sdtyping}.
% In this section, we focus on chosen parts of our system
% through a set of selected rules.
%
We discuss judgments for typing, kinding, and constraint entailment:
\begin{description}
\item[$\inferS{C}{\E}{e}{\tau}$ --]
  Expression $e$ has type $\tau$ in environment $\E$ under constraints $C$.
\item[$\inferSK{C}{\E}{\tau}{k}$ --]
  Type $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
\item[$\entail{D}{C}$ --] Constraint $D$ entails constraint $C$.
\end{description}

@@

\affe uses of kinds and constrained types to indicate
linear and affine types.
% Before explaining the details of our type system, we
% give a quick presentation of the kinds and constraints.
A kind $k$ is either a kind variable, $\kvar$, or a constant.
Each constant is composed of a ``quality'' and a ``level''.
The quality describes the use pattern of the type. It can be
unrestricted $\kun$, affine $\kaff$, or linear $\klin$. The level  $n \in \Nat \cup \{\infty\}$
describes the nested regions in which the value can be used.
Level $0$ refers to the top-level scope outside any region; we often elide it
and write $\kaff$ for $\kaff_0$. Level
$\infty$ refers to an empty region that is infinitely nested.
For instance, the constraint $\Cleq{\kvar}{\kun_\infty}$ indicates that
$\kvar$ must be unrestricted, but can be local to a region.

Kinds form a lattice with an ordering that implements two principles
(see \cref{sdtyp:lattice}). First, unrestricted values may be used
where affine ones are expected and affine ones are less restricted
than linear one as reflected in \TirName{Lat-UA} and \TirName{Lat-AL}.
Second, values usable at level $n$ may be used at any more deeply
nested level $n'$ (where $n\le\infty$ for any $n$) as defined in the \TirName{Lat-X-Level} axioms.

Constraints are conjunctions of inequality constraints over this kind
lattice. In most uses, a constraint gives an upper or lower bound for
a kind variable or it relates two kind variables.

\begin{figure*}[tp]
  \input{infer/lattice}
  \caption{Lattice ordering -- $k \lk_\Lat k'$}
  \label{sdtyp:lattice}
\end{figure*}
